---
title: "Getting started with foundry10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with foundry10}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
library(foundry10)
library(ggplot2)
```

## Setup

Install from GitHub (once) and load as normal:

```r
# install.packages("remotes")
remotes::install_github("kgweisman/foundry10")
library(foundry10)
library(ggplot2)
```

The only required dependency is **ggplot2**. Install **systemfonts** for
automatic Helvetica Neue font detection:

```r
install.packages("systemfonts")
```

---

## The main theme

Add `theme_foundry10()` to any ggplot. The typical pattern is:

```r
ggplot(data, aes(x, y, color = group)) +
  geom_point() +
  scale_color_foundry10_d() +
  theme_foundry10() +
  labs(title = "Title", subtitle = "Subtitle", caption = "Source: ...")
```

`theme_foundry10()` accepts three arguments:

| Argument | Default | Options |
|---|---|---|
| `base_size` | `12` | Any number (pt). Use 14-16 for slides. |
| `grid` | `TRUE` | `FALSE` removes horizontal grid lines. |
| `strip_fill` | `"white"` | `"orange"`, `"cyan"`, `"black"`, `"white"` |

Quick preview:

```{r preview}
preview_theme_foundry10()
```

---

## Choosing a color scale

| Data type | Function | Notes |
|---|---|---|
| Categorical -- no valence | `scale_fill_foundry10_d()` | Up to 6 categories |
| Continuous -- negative/concerning | `scale_fill_foundry10_seq()` | Orange ramp (default) |
| Continuous -- positive/favorable | `scale_fill_foundry10_seq(tint = "cyan")` | Cyan ramp |
| Continuous diverging | `scale_fill_foundry10_div()` | Orange -> gray -> cyan |
| Ordered categories, 2-4 levels | `scale_fill_foundry10_seq_d(n)` | Fixed brand tints |
| Likert / ordered with midpoint, 2-9 levels | `scale_fill_foundry10_div_d(n)` | Orange -> gray -> cyan |

All scale functions have `fill` and `color` variants and `scale_colour_`
aliases. Most accept `direction = -1` to reverse the ramp.

### Discrete (qualitative)

```{r discrete}
mpg6 <- mpg[mpg$class != "minivan", ]
ggplot(mpg6, aes(displ, hwy, color = class)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_foundry10_d() +
  theme_foundry10() +
  labs(title    = "Engine displacement vs highway MPG",
       subtitle = "By vehicle class",
       x = "Displacement (L)", y = "Highway MPG")
```

### Sequential

```{r sequential}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_foundry10_seq() +
  theme_foundry10(grid = FALSE) +
  labs(title = "Old Faithful eruption density") +
  guides(fill = guide_colorbar(barwidth = 20))
```

### Diverging (continuous)

```{r diverging}
set.seed(42)
regions <- c("Northeast", "Southeast", "Midwest", "West", "Southwest")
df_div  <- expand.grid(region = regions,
                       month  = factor(month.abb, levels = month.abb))
df_div$variance <- round(rnorm(nrow(df_div), mean = 0, sd = 12), 1)

ggplot(df_div, aes(x = month, y = region, fill = variance)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_foundry10_div(limits = c(-30, 30),
                           name   = "vs. prior year (%)") +
  theme_foundry10(grid = FALSE) +
  labs(title    = "Regional sales variance by month",
       subtitle = "Percentage change vs. prior year",
       x = NULL, y = NULL) +
  guides(fill = guide_colorbar(barwidth = 15))
```

### Diverging discrete (Likert)

```{r likert}
set.seed(1)
df_likert <- data.frame(
  response = factor(
    sample(c("Strongly disagree", "Disagree", "Neutral",
             "Agree", "Strongly agree"), 200, replace = TRUE),
    levels = c("Strongly disagree", "Disagree", "Neutral",
               "Agree", "Strongly agree")
  )
)
ggplot(df_likert, aes(x = response, fill = response)) +
  geom_bar() +
  scale_fill_foundry10_div_d(n = 5) +
  theme_foundry10() +
  labs(title = "Survey responses", x = NULL, y = "Count") +
  theme(legend.position = "none")
```

---

## Data viz color semantics

Per the **foundry10 Visual Style Guide**:

- **Orange** = negative, bad, concerning, or below-target values
- **Cyan** = positive, good, favorable, or above-target values
- **Gray** = neutral, unimportant, or background elements

> The qualitative palette (`scale_fill_foundry10_d`) is for categorical data
> with no positive/negative meaning. Avoid assigning orange (slot 1) to a
> "yes", "agree", or otherwise positive category.

---

## Utility functions

```r
foundry10_color("orange_foundry10")    # "#D14E1D"
foundry10_color("cyan_cool")           # "#007079"

foundry10_palette("primary")           # named vector of 4 primary colors
foundry10_palette("all")               # all brand colors

preview_foundry10_palette("secondary") # renders a color swatch
```
